---
title: "archeoViz Vignette"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{archeoViz Vignette}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```




`archeoViz` is a packaged R Shiny application for the *visualisation*, *exploration*, and web *communication* of archaeological spatial data. It includes interactive 3D and 2D *visualisations*, can generate *cross sections* and *maps* of the remains, can run basic *spatial statistics* methods (convex hull, regression surfaces, 2D kernel density estimation), and display an interactive *timeline* of an excavation. `archeoViz` can be used locally or deployed on a server, either by allowing the user to load data through the interface or by running the app with a specific data set. The app interface is available in English, French, Italian, and Portuguese.


- [**Installation**](#installation)
  - [Local use](#local-use)
  - [Remote use](#remote-use)
  - [Demonstration](#demonstration)
- [**Community guidelines**](#community-guidelines)
  - [Reporting bugs](#reporting-bugs)
  - [Suggesting changes](#suggesting-changes)
- [**Use**](#use)
  - [Data input](#data-input)
    - [Loading files through the Input data tab](#loading-files-through-the-input-data-tab)
    - [Random data](#random-data)
    - [Through function parameters](#through-function-parameters)
    - [Through an URL](#through-an-url)
  - [Data sub-setting](#data-sub-setting)
    - [Location mode](#location-mode)
    - [Objects category](#objects-category)
    - [Data subgroups](#data-subgroups)    
    - [Objects selection](#objects-selection)
  - [Interactive visualisation](#interactive-visualisation)
  - [Graphical outputs](#graphical-outputs)
  - [Spatial statistics](#spatial-statistics)  
    - [Regression surfaces](#regression-surfaces)
    - [Convex hulls](#convex-hulls)
    - [2D kernel density](#2d-kernel-density)
- [**Reproducibility**](#reproducibility)
- [**Advanced parameters**](#advanced-parameters)
    - [Square grid](#square-grid)
    - [Parameter presetting](#parameter-presetting)
    - [Reactive plot display](#reactive-plot-display)
    - [URL parameters](#url-parameters)
- [**Acknowledgments**](#acknowledgments)
- [**References**](#references)
 
# Installation

`archeoViz` can be used in two ways:

* locally, on the user's machine
* remotely, after deploying the app on a distant server

## Local use


The package can be installed from CRAN with:

```r
install.packages("archeoViz")
```

The development version is available on *GitHub* and can be installed with:

```r
# install.packages("devtools")
devtools::install_github("sebastien-plutniak/archeoviz")
```

Then, load the package and launch the app with:

```r
library(archeoViz)
archeoViz()
```

## Remote use

To deploy `archeoViz` on your Shiny server, first download and unzip the package:

```{r, eval=FALSE}
# set the working directory on your shiny server:
setwd(dir = "/some/path/")
# download the package:
download.file(
  url = "https://github.com/sebastien-plutniak/archeoviz/archive/master.zip",
  destfile = "archeoviz.zip")
# unzip it:
unzip(zipfile = "archeoviz.zip")
```

Then, go to `https://<your-shiny-server>/archeoviz-main`.

To set the app with your data and preferences, edit the app.R file, located at the root of the directory:

```{r, eval=FALSE}
archeoViz(objects.df = NULL,   # data.frame with data about the objects
          refits.df = NULL,    # optional data.frame for refitting data
          timeline.df = NULL,  # optional data.frame for the excavation timeline
          title = NULL,        # title of the site / data set
          home.text = NULL,    # HTML content to display on the home page
          lang = "en"          # interface language ("en": English, "fr": French, "it": Italian, "pt": Portuguese)
          set.theme = "cosmo") # graphic theme for the Shiny interface
```

The possible values for the `set.theme` parameter are illustrated on [this page](https://rstudio.github.io/shinythemes/).
The language of the application can be set with the `lang` parameter.

## Demonstration

Demonstration instances of the application are deployed on the *Huma Num* Shiny server:

* [`archeoViz` in English](https://analytics.huma-num.fr/archeoviz/en).
* [`archeoViz` in French](https://analytics.huma-num.fr/archeoviz/fr).
* [`archeoViz` in Italian](https://analytics.huma-num.fr/archeoviz/it).
* [`archeoViz` in Portuguese](https://analytics.huma-num.fr/archeoviz/pt).

Real use cases are presented on the [*archeoViz Portal*](https://analytics.huma-num.fr/archeoviz/home).


# Community guidelines

## Reporting bugs
If you encounter a bug, please fill an [issue](https://github.com/sebastien-plutniak/archeoviz/issues) with all the details needed to reproduce it.

## Suggesting changes
Suggestions of changes to `archeoViz` are welcome.
These requests may concern additional functions, changes to documentation, additional examples, new features, etc.
They can be made by filling an [issue](https://github.com/sebastien-plutniak/archeoviz/issues) and, even better, using pull requests and the [GitHub Fork and Pull model](https://docs.github.com/articles/about-pull-requests).


# Use

Having archaeological remains from a given site, `archeoViz` is designed to lower the technical barriers to fulfil three objectives:

* basic spatial exploration and generation of simple graphical reports;
* fast pre-publication of archaeological data, intended for the scientific community;
* fast deployment of a display and communication tool intended for a broader audience.

In addition, `archeoViz` is a suitable pedagogical resource for teaching spatial analysis in archaeology, data structuring, open science, and reproducibile workflow.

N.B.: consequently, `archeoViz` is not intended to replace more sophisticated analysis tools (e.g., GIS, statistical packages, etc.)

## Data input

There are four ways to input data in `archeoViz`:

1. uploading tables in the “Input data” tab,
2. using randomly generated data from the “Input data” tab;
3. set the `archeoViz` main function's parameters before running the application.
4. through the URL parameters, when using an online instance of `archeoViz`

### Loading files through the Input data tab

Tables for three types of data can be uploaded from the “Input data” tab:

* an “objects” table (mandatory), with data about the objects;
* a “refits” table (optional), with data about the refitting objects;
* a “timeline” table (optional), with data about when each square of the site was excavated.

The tables must be CSV files with the first row used containing the columns' labels (the separator can be set). Contents in HTML are allowed. This makes it possible, in particular, to add links to external resources (e.g., the permanent identifier of the object's in other databases, or of the concepts used to describe the object, etc.).

#### Objects table

A row describes a single object with the following mandatory fields:

 * **id**: *alphanumerical value*, unique identifier of the object 
* **xmin**: *numerical value*, coordinate of the object on the X axis (in cm)
* **ymin**: *numerical value*, coordinate of the object on the Y axis (in cm)
* **zmin**: *numerical value*, coordinate of the object on the Z axis (positive depth value in cm)
* **layer**: *alphanumerical value*, identifier of the object's layer
* **object_type**: *alphanumerical value*, category of the object

In addition, optional fields are possible, including:

* **square_x**: *alphanumerical value*, identifier of the square on the X axis
* **square_y**: *alphanumerical value*, identifier of the square on the Y axis
* **year** : *numerical value*, year when the object was excavated
* **xmax**: *numerical value*, when the X location of the object is included in a range of X coordinates
* **ymax**: *numerical value*, when the Y location of the object is included in a range of Y coordinates
* **zmax**: *numerical value*, when the Z location of the object is included in a range of Z coordinates
* **object_edit**: unlimited number of additional variables describing the object (field names must start with `object_` and have different suffixes)
 
The labels of the squares of the grid:

* are ordered alpha-numerically;
* are not displayed to avoid erroneous displays, if the number of labels does not correspond exactly to the total number of 100 cm squares that can be defined in the range of minimum and maximum coordinates contained in the xmin and ymin variables;
* can be completed with the `add.x.square.labels` and `add.y.square.labels` parameters of the `archeoViz()` function in order to add the missing labels (on the X and Y axes of the grid, respectively).


### Random data

For demonstration purposes using randomly generated data is made possible. To activate this feature, set the slider in “Input data” to a value higher than 0 (setting the value back to 0 deactivates the feature). An “objects” data set, a “refits” data set, and a “timeline” data set are generated, making it possible to test all the `archeoViz` functionalities.

### Through function parameters

`archeoViz`'s launching function (`archeoViz()`) can be run without parameter

```{r, eval=FALSE}
archeoViz()
```

or by using the `objects.df`, `refits.df`, or `timeline.df` parameters to input data.frames about the archaeological objects, refitting relationships between these objects, and the chronology of the excavation, respectively.

```{r, eval=FALSE}
archeoViz(objects.df = NULL,  # data.frame with data about the objects
          refits.df = NULL,   # data.frame for refitting objects
          timeline.df = NULL) # optional data.frame for the excavation timeline
```

### Through an URL

The URL of an online instance of `archeoViz`, for example, https://analytics.huma-num.fr/archeoviz/en/, can include a `?objects.df=` parameter, whose value is the URL of a CSV file observing the `archeoViz` format.

## Data sub-setting

Once data are loaded, a sub-selection of the data set can be done in the left side menu. Several parameters are possible: the type of location recording, the category of the objects, and the way to group the data. 

### Location mode

The location of archaeological objects may have been recorded in different, more or less precise ways: on the one hand, exactly with points (located by a triplet of x, y and z coordinates) and, on the other hand, more or less imprecisely with lines, planes, or within volumes (with different sets of x, y and/or z value pairs).
In `archeoViz`, a distinction is made between exact location (points) and other types of vague location (located on lines, planes, and volumes).
Both types of locations (exact and vague) can be displayed, and location uncertainties can also be visualised as lines, planes and volumes. This last option is resource intensive, and using it with too much data can significantly slow down the application.


### Objects category

Sub-sets can be defined by object categories, using the “variable” and “values” fields.
Once one of the “object_type” (or other possible “object_” variables) is selected, its values appear below and can be selected using the tick boxes.
The selection must be validated by clicking on the “Validate” button.
This selection determines the data that will be displayed in the plots and tables.

### Data subgroups

The data can be grouped in two ways: either by layer or by the selected “object_” variable.
This option determines the colours of the points in the 3D and 2D plots and the subsets when computing surfaces and convex hulls.

### Objects selection

In the “3D plot”, “Map”, “Section X”, and “Section Y” tabs, clicking on a point displays information about that point in the table below the plot.

## Interactive visualisation

The plots in the “3D plot”, “Map”, “Section X”, and “Section Y” tabs are generated using the [`plotly`](https://CRAN.R-project.org/package=plotly) library. All the plots are dynamic and include a menu bar above the plot with several options (generating an image file, zooming, moving the view, etc). See details on the
[`plotly` website](http://plotly.github.io/getting-to-know-the-plotly-modebar/).

Clicking on a legend's item modifies the display:

* a simple click on an item activates/deactivates its display;
* a double click on an item displays this item only (another double click cancels it).

This feature makes it possible to choose which layers are shown.
In addition, the size of the points can be set and whether the refitting relationships must be represented or not.

## Graphical outputs

Several graphical outputs can be generated in `archeoViz`.

* The plots in The plots in the “3D plot”, “Map”, “Section X”, and “Section Y” can be exported:
	* in SVG format (by clicking on the “camera” icon in the menu bar above the plot),
	* in an interactive HTML format, by clicking on the "Export" button.
* The plan of the excavation chronology can be exported in SVG format by clicking on the “Download” button.

## Reffitings

Refittings are usually recorded by archaeologists in two ways:

1. by sets of refitting objects: using a two columns table, where a row corresponds to an **object**. The first column stores the object's unique id and the second column stores the id of the set of refitting objects this object belongs to.
2. by refitting relationships: using a two columns table, where a row corresponds to a **relationship**. The first column stores the first object's unique id and the second column stores the second object's unique id.

Although the second data structure is more accurate, the first is more commonly used. 

`archeoViz` processes the two data structures in different ways:

* sets of refitting objects must be described using a specific column in the `objects.df` table (e.g. `object_refits`) and are represented by the color of points in the plots (as for any other variable);
* refitting relationships must be described using the `refits.df` table and are visualised as segments connecting the refitting objects in the plots.

## Spatial statistics

`archeoViz` includes some spatial analysis functionalities, intended for basic and exploratory use.

### Regression surfaces

In the “3D plot” tab, clicking on “Compute surfaces” and “Validate” displays the regression surface associated with each layer (with at least 100 points). The surfaces are computed using the generalized additive model implemented in the [`mgcv`](https://CRAN.R-project.org/package=mgcv) package.

### Convex hulls

In the “3D plot” tab, clicking on “Compute hulls” and “Validate” displays the convex hull associated with each layer (with at least 20 points). The convex hulls are computed using the  [`cxhull`](https://CRAN.R-project.org/package=cxhull) package.

### 2D kernel density

In the “Map” tab, ticking the “Compute density” box and clicking on “Validate” generates a map with contour lines showing the points' density.
Density can be computed for all the points together or by layer (with at least 30 points).
The 2D kernel density is computed with the `kde2d` function of the [`MASS`](https://CRAN.R-project.org/package=MASS) package (through [`ggplot2`](https://CRAN.R-project.org/package=ggplot2)).


## Reproducibility

`archeoViz` is, by definition, an interactive application. However, several features guarantee the reproducibility and communicability of the result of interactions with the application.

* The 3D visualisation can be exported in an interactive HTML standalone format, considering the data selection made by the user.
* In the “Reproducibility” tab, an R command is dynamically generated, considering the current application settings made by the user.
* In a more advanced use, the URL parameters makes it possible to set an online instance of the application parameters of interest and to communicate it by sending the URL.

## Advanced parameters

The `archeoViz()` function can be set with multiple optional parameters, related to:

* the input data  (already detailed [above](#through-function-parameters)),
* the contents of the home page (already detailed [above](#deployed-use)),
* the [square grid](#square-grid),
* the [presetting](#parameters-presetting) of the parameters that can be set through the application's interface,
* the [reactive behavior](#reactive-plot-display) of the application regarding the generation of plots.
* the [HTML export](#html-export).
* the [URL parameters](#url-parameters).

```{r, eval=FALSE}
archeoViz(objects.df=NULL, refits.df=NULL, timeline.df=NULL,
          title=NULL, home.text=NULL, lang="en", set.theme="cosmo",
          square.size = 100, reverse.axis.values = NULL, reverse.square.names = NULL,
          add.x.square.labels = NULL, add.y.square.labels = NULL,
          class.variable = NULL, class.values = NULL,
          default.group = "by.layer", location.mode = NULL,
          map.z.val = NULL, map.density = "no", map.refits = NULL,
          plot3d.ratio = 1, plot3d.hulls = NULL, plot3d.surfaces = NULL, plot3d.refits = NULL,
          sectionX.x.val = NULL, sectionX.y.val = NULL, sectionX.refits = NULL, 
          sectionY.x.val = NULL, sectionY.y.val = NULL, sectionY.refits = NULL,
          camera.center = c(0, 0, 0), camera.eye = c(1.25, 1.25, 1.25),
          run.plots = FALSE, html.export = TRUE
          )
```

### Square grid

```{r, eval=FALSE}
archeoViz(square.size = 100,
          reverse.axis.values = NULL, reverse.square.names = NULL,
          add.x.square.labels = NULL, add.y.square.labels = NULL
          )
```

* **square.size**: numerical. Size (width and height) in centimeter of the squares in the grid system. Default value is 100 cm.
* **reverse.axis.values**: character. Name of the axis or axes to be reversed (any combination of “x”, “y”, “z”).
* **reverse.square.names**: character. Name of the axis or axes for which to reverse the order of the square labels (any combination of “x”, “y”, “z”).
* **add.x.square.labels**: character. Additional square labels for the “x” axis. 
* **add.y.square.labels**: character. Additional square labels for the “y” axis.

### Parameter presetting

```{r, eval=FALSE}
archeoViz(class.variable = NULL, class.values = NULL,
		  default.group = "by.layer", location.mode = NULL,
		  map.z.val = NULL, map.density = "no", map.refits = NULL,
		  plot3d.hulls = NULL, plot3d.surfaces = NULL, plot3d.refits = NULL,
		  sectionX.x.val = NULL, sectionX.y.val = NULL, sectionX.refits = NULL, 
		  sectionY.x.val = NULL, sectionY.y.val = NULL, sectionY.refits = NULL,
		  camera.center = NULL, camera.eye = NULL
          )
```

* **class.variable**: character. At the launch of the app, name of the variable to preselect.
* **class.values**: character vector. At the launch of the app, names of the values to preselect.
* **default.group**: character. At the launch of the app, preselection of the variable used to group data (one of “by.layer” or “by.variable”).
* **location.mode**: character. At the launch of the app, preselection of the location method (one of “exact”, “fuzzy”, “show.uncertainty”).
* **map.z.val**: numerical. Minimal and maximal Z coordinates values to display in the map plot.
* **map.density**: character. At the launch of the app, whether to compute and show density contours in the map plot (one of “no”, “overall”, “by.variable”).
* **map.refits**: TRUE or FALSE. Whether to show refits in the map plot.
* **plot3d.hulls**: TRUE or FALSE. At the launch of the app, whether to compute and show convex hulls in the 3D plot.
* **plot3d.surfaces**: TRUE or FALSE. At the launch of the app, whether to compute and show regression in the 3D plot.
* **plot3d.refits**: TRUE or FALSE. At the launch of the app, whether to show refits on the 3D section plot.
* **sectionX.x.val**: numerical. At the launch of the app, minimal and maximal X coordinates values to display in the X section plot.
* **sectionX.y.val**: numerical. At the launch of the app, minimal and maximal Y coordinates values to display in the X section plot.
* **sectionX.refits**: TRUE or FALSE. At the launch of the app, whether to show refits in the X section plot.
* **sectionY.x.val**: numerical. At the launch of the app, minimal and maximal X coordinates values to display in the Y section plot.
* **sectionY.y.val**: numerical. At the launch of the app, minimal and maximal Y coordinates values to display in the Y section plot.
* **sectionY.refits**: TRUE or FALSE. At the launch of the app, whether to show refits in the Y section plot.
* **camera.center**: numerical. In 3D plot, coordinates of the point to which the camera looks at (default values: x=0, y=0, z=0).
* **camera.eye**: numerical. In 3D plot, coordinates of the camera's position (default values: x=1.25, y=1.25, z=1.25).


### Reactive plot display

```{r, eval=FALSE}
archeoViz(run.plots = FALSE)
```

* **run.plots**: TRUE or FALSE. Whether to immediately compute and show plots (without requiring the user to click on the buttons in the interface).

### HTML export

* **html.export** : TRUE or FALSE.  Whether or not to allow figures to be exported as interactive HTML widgets.

### URL parameters

An instance of `archeoViz` deployed online on a server can be set with URL parameters.
Supported parameters include:

* `objects.df`, `refits.df`, `timeline.df`
* `title`, `home.text`
* `reverse.axis.values`, `reverse.square.names`
* `square.size`
* `add.x.square.labels`, `add.y.square.labels`
* `class.variable`, `class.values`
* `default.group`
* `location.mode`
* `map.density`, `map.refits`
* `plot3d.hulls`, `plot3d.surfaces`, `plot3d.refits`
* `sectionX.refits`
* `sectionY.refits`
* `run.plots`


(The following parameters are not supported in the current version: `map.z.val`, `sectionX.x.val`, `sectionX.y.val`, `sectionY.x.val`, `sectionY.y.val`, `lang`, `set.theme`, `camera.center`, `camera.eye`, `html.export`.)

The parameters must be written using the URL syntax (?param1=value&param2=value2) and have the same type of values than when used in the R interface.
For example, the following URL launches an `archeoViz` instance using the [Bilzingsleben](https://zenodo.org/record/8003880) dataset:

https://analytics.huma-num.fr/archeoviz/en/?objects.df=https://zenodo.org/record/8003880/files/bilzingsleben.csv

This URL does the same, but also includes the refitting table (parameter `&refits.df=`) and set the activate the display of the refitting relationships in the 3D and map plots:

https://analytics.huma-num.fr/archeoviz/en/?map.refits=TRUE&plot3d.refits=TRUE&objects.df=https://zenodo.org/record/8003880/files/bilzingsleben.csv&refits.df=https://zenodo.org/record/8003880/files/bilzingsleben-antlers-refits.csv

The following URL launches the Bilzingsleben dataset, pre-setting the app to: 

1. groups the points by variable (parameter `default.group`, with walue `by.variable` instead of `by.layer`)
2. selects only the "Antlers" (parameter `class.values`)
3. redefines the size of the square grid (parameter `square.size`, 500 cm instead of the 100 cm default value)
4. enable the immediate display of the plots (parameter `run.plots`)
5. modifies the title of the page (parameter `title`)
6. modifies the content of the home page with basic html contents (parameter `home.txt`)

[https://analytics.huma-num.fr/archeoviz/en/?default.group=by.variable&class.values=Antler&square.size=500&run.plots=TRUE&title=Antlers%20at%20Bilzingsleben&home.text=Many%20<b>antlers</b>&objects.df=https://zenodo.org/record/8003880/files/bilzingsleben.csv](https://analytics.huma-num.fr/archeoviz/en/?default.group=by.variable&class.values=Antler&square.size=500&run.plots=TRUE&title=Anters%20at%20Bilzingsleben&home.text=Many%20<b>antlers</b>&objects.df=https://zenodo.org/record/8003880/files/bilzingsleben.csv)
 
 
Note that the parameters `reverse.axis.values`, `reverse.square.names`, `add.x.square.labels`, `add.y.square.labels`, `location.mode`, and `class.values`, which accept simple or multiple values in the R interface (e.g. c("value1", "value2")) only accept one value when set as URL parameters (this is a restriction due to the URL syntax).


# Acknowledgments

The `archeoViz` application and package is developed and maintained by Sébastien Plutniak. Arthur Coulon, Solène Denis, Olivier Marlet, and Thomas Perrin tested and supported the project in its early stage. Renata Araujo and Sara Giardino translated the application into Portuguese and Italian, respectively.

# References

* Plutniak, Sébastien, Renata Araujo, Sara Giardino. 2023. “archeoViz. Visualisation, Exploration, and Web Communication of Archaeological Excavation Data”. v1.1.2, DOI: [10.5281/zenodo.7682227](https://doi.org/10.5281/zenodo.8000631).
* Plutniak, Sébastien. 2023. “[Visualiser et explorer la distribution spatiale du mobilier archéologique: l'application archeoViz et son portail web](https://www.prehistoire.org/offres/doc_inline_src/515/0-BSPF_2023_1_2e_partie_Correspondance_PLUTNIAK.pdf)”. *Bulletin de la Société préhistorique française*, 120(1), p. 70-74.
